Draft_Application_Stochastic_Volatility

### 2021-08-11 Real Dataset SV.Rmd
Riferimenti intratesto

In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data. However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques. Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.\
This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them. In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model. Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques. In section \textit{fill with the number of the section} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \textit{fill with the number of the section} for synthetic data.

\section{A review of Stochastic Volatility models}

In section \textit{fill with the number of the section} a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset. Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows, 
\begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}

@@

\section{Data Description}

@@

As regards the series that we will employ in section \textit{fill with the number of the section}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.\footnote{Such class of measures will be better presented in section \textit{fill with the number of the section}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.} We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals. Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021. 
The plots below represent the realized volatility in the period of interest for all the three indices.\footnote{In the graph, the series $rv5$ is rescaled as specified in the upcoming section: we took the square root of each value, which is a daily variance measure, and multiplied it by 100.} As usual, they show a rather similar pattern across all three indices, identifying peaks corresponding to periods of higher uncertainty (in particular referring to the 2020 Covid-19 Crisis outbreak and its impact over time). Once more, there is great similarity between the series for the DJI and the one for the S&P500, while the series of the STOXX50E appears slightly different, in terms of peaks.

@@

\section{Comparing the filters}

In this section the different particle filter techniques we discussed in earlier sections will be employed to approximate the filtering distribution that springs from a simple SV model applied to a real dataset. First, we provide a qualitative comparison between the filters, basing on visual analysis of the (approximated) filtering distributions. Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases. In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.\
Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model. Indeed, the Liu and West filter actually approximates the distribution of a slightly different SV model, one that assumes an additional hierarchical level,
\begin{align*}
y_t|x_t & \sim N(0,e^{x_t})\\
x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
x_0 & \sim N(0,100)\\
\alpha & \sim N(\gamma,\zeta)\\
\beta & \sim N(\pi,\phi)\\
\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
\end{align*}
where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\

@@

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling algorithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms. Finally, note that the Auxiliary Particle Filter employed here does not coincide with that of the section with synthetic data, which employed a different kernel implied by optimality results.}

@@

Using the APF as the benchmark does not deliver a different picture than the one we previously had. Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at each threshold (though very slightly so for the higher ones). Its performance ameliorates dramatically at N=1000, while it still shows some sizable improvements even when passing from N=5000 to N=10000, almost halving its errors. Moreover, note that the mere fact that at N=10000 the Guided and the Bootstrap particle filters approximate better the APF benchmark suggests that the APF still has some margins for relative improvement even at high levels of N, pointing at higher computational costs than the competing filters: trivially, since at N=50000 its RMS and MA difference with respect to this benchmark must be zero, the APF overtakes the Guided and the Bootstrap PFs at N$>>10000$.\

@@

\section{Realised volatility as benchmark}

In section XX we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters. However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality. In this section, we will employ realised volatility as one such metric.\
A realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices presented in the earlier sections. As we previously mentioned, a transformation is needed as to compare filtered states and realised volatility, namely one turning both into estimates of the standard deviation of the percentage log returns. For each $t$, $rv5_t$ is then turned into $rv5_t^{0.5}*100$, while $e^{x_t/2}$ (where $x_t$ is the filtered state) is computed by taking the mean of the generated particles and then applying the $exp(\cdot /2)$ transformation.\
Throughout this section, we will deem as the best performing algorithm that coming closest to the proxy, either "literally" in a graphical framework or in a Root Mean Squared Error or Mean Absolute Error sense. Clearly enough, this approach builds on the assumption that realised variance is indeed a good proxy for latent volatility in real financial series.\

@@

In order to evaluate the different filters according to the evaluation tool represented by the realized volatility, we start once more from the visual inspection of the plots. For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} As compared to the visual analysis we conducted in section \textit{fill with the number of the section}, we can now use such a proxy to determine the relative performance of the competing filters.\

@@

The above intuitions are confirmed by looking at Table XX, where the RMSE and MAE are shown for each filter (run with a number of particles equal to 10000), considering the realised volatility as benchmark. Indeed, for any index, most of the filters give rather similar results in terms of RMSE and MAE. For any index and for any measure, the worst performing is the SIS algorithm, while the best performing is the Liu and West particle filter. Although the latter algorithm was outperformed in the analysis we conducted on synthetic data, in fact we expected such reversal of fortunes: intuitively, while the rest of the field is built on a model with fixed guesses for the unknown parameters, the Liu and West filters enjoys more flexibility thanks to the higher variability implied by its additional hierarchical level and by its parameter learning, two features that should both pay off in real data analyses.\

@@

Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices. We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF. Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a 0.3794359 RMSE and a 0.2687336 MAE. In words, worse approximations of the target distribution of the LWF are actually best at replicating the proxy for the SP500. A possible interpretation goes, once again, as that reported in the main text.}\
In the case of the EuroStoxx50 index, while the Bootstrap filter slightly increases both RMSE and MAE when N is raised to 50000, the APF shows mild improvements in its MAE performance, while the RMSE criterion points in the opposite direction. The criteria disagree also in the case of the LW filter, as the MAE mildly prefers the performance of N=10000, while the RMSE detects minor improvements for N=50000.

>>>>

In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data. However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques. Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.\
This part of the analysis is structured as follows: section \ref{Literary review} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \ref{Data description} describes the data which we employ and the transformations that were applied to them. In section \ref{Comparing the filters} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model. Section \ref{Forecast comparison} features a forecast comparison between a SV specification and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques. In section \ref{Realised volatility} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \ref{Simulated SV data} for synthetic data.

\section{A review of Stochastic Volatility models}\label{Literary review}

In section \ref{Simulated SV data} a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset. Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows, 
\begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}

@@

\section{Data Description}\label{Data description}

@@

As regards the series that we will employ in section \ref{Realised volatility}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.\footnote{Such class of measures will be better presented in section \ref{Realised volatility}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.} We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals. Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021. 
The plots below represent the realized volatility in the period of interest for all the three indices.\footnote{In the graph, the series $rv5$ is rescaled as specified in the upcoming section: we took the square root of each value, which is a daily variance measure, and multiplied it by 100.} As usual, they show a rather similar pattern across all three indices, identifying peaks corresponding to periods of higher uncertainty (in particular referring to the 2020 Covid-19 Crisis outbreak and its impact over time). Once more, there is great similarity between the series for the DJI and the one for the S&P500, while the series of the STOXX50E appears slightly different, in terms of peaks.

@@

\section{Comparing the filters}\label{Comparing the filters}

In this section the different particle filter techniques we discussed in earlier sections will be employed to approximate the filtering distribution that springs from a simple SV model applied to a real dataset. First, we provide a qualitative comparison between the filters, basing on visual analysis of the (approximated) filtering distributions. Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases. In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.\
Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model. Indeed, the Liu and West filter actually approximates the distribution of a slightly different SV model, one that assumes an additional hierarchical level,
\begin{align*}
y_t|x_t & \sim N(0,e^{x_t})\\
x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
x_0 & \sim N(0,100)\\
\alpha & \sim N(\gamma,\zeta)\\
\beta & \sim N(\pi,\phi)\\
\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
\end{align*}
where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part.}. The initial state is drawn from a normal distribution with commonly used parameters for percentage log returns, as reported by Kastner (2019).\
For simplicity, in $$y_t|x_t \sim N(0,e^{x_t})$$ $$x_t|x_{t-1},\alpha,\beta,\tau^2 \sim N(\alpha+\beta x_{t-1},\tau^2)$$ we set $\alpha=0$, a value that in general, as we argued earlier, also has an empirical justification for percentage log returns. We choose $\beta=0.99$, as one of the basic underlying ideas of SV models is considerable persistence in the volatility. Finally, we set $\tau^2=0.05$.\
In the specification used by the LW filter, we have $\gamma=0$ and $\pi=0.99$, as to center the priors on the point guesses of the simpler model; $\zeta$ and $\phi$ are both set to $0.01$, then allowing a 0.1 standard deviation on both $\alpha$ and $\gamma$. For $\tau^2$, we choose an Inverse Gamma distribution whose expected value coincides with the point guess $0.05$, by setting $\nu=5$ and $\lambda=0.05*3/5$.\

@@

We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark. In the table below we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles. Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below. Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling algorithm. By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never). In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms. Finally, note that the Auxiliary Particle Filter employed here does not coincide with that of the section with synthetic data, which employed a different kernel implied by optimality results.}

@@

Using the APF as the benchmark does not deliver a different picture than the one we previously had. Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at each threshold (though very slightly so for the higher ones). Its performance ameliorates dramatically at N=1000, while it still shows some sizable improvements even when passing from N=5000 to N=10000, almost halving its errors. Moreover, note that the mere fact that at N=10000 the Guided and the Bootstrap particle filters approximate better the APF benchmark suggests that the APF still has some margins for relative improvement even at high levels of N, pointing at higher computational costs than the competing filters: trivially, since at N=50000 its RMS and MA difference with respect to this benchmark must be zero, the APF overtakes the Guided and the Bootstrap PFs at N $>>10000$.\

@@

\section{Realised volatility as benchmark}\label{Realised volatility}

In section \ref{Comparing the filters} we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters. However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality. In this section, we will employ realised volatility as one such metric.\
A realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices presented in the earlier sections. As we previously mentioned, a transformation is needed as to compare filtered states and realised volatility, namely one turning both into estimates of the standard deviation of the percentage log returns. For each $t$, $rv5_t$ is then turned into $rv5_t^{0.5}*100$, while $e^{x_t/2}$ (where $x_t$ is the filtered state) is computed by taking the mean of the generated particles and then applying the $exp(\cdot /2)$ transformation.\
Throughout this section, we will deem as the best performing algorithm that coming closest to the proxy, either "literally" in a graphical framework or in a Root Mean Squared Error or Mean Absolute Error sense. Clearly enough, this approach builds on the assumption that realised variance is indeed a good proxy for latent volatility in real financial series.\

@@

In order to evaluate the different filters according to the evaluation tool represented by the realized volatility, we start once more from the visual inspection of the plots. For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} As compared to the visual analysis we conducted in section \ref{Comparing the filters}, we can now use such a proxy to determine the relative performance of the competing filters.\

@@

The above intuitions are confirmed by looking at the table below, where the RMSE and MAE are shown for each filter (run with a number of particles equal to 10000), considering the realised volatility as benchmark. Indeed, for any index, most of the filters give rather similar results in terms of RMSE and MAE. For any index and for any measure, the worst performing is the SIS algorithm, while the best performing is the Liu and West particle filter. Although the latter algorithm was outperformed in the analysis we conducted on synthetic data, in fact we expected such reversal of fortunes: intuitively, while the rest of the field is built on a model with fixed guesses for the unknown parameters, the Liu and West filters enjoys more flexibility thanks to the higher variability implied by its additional hierarchical level and by its parameter learning, two features that should both pay off in real data analyses.\

@@

Finally, we can see from the table below the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices. We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section \ref{Comparing the filters}; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF. Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a 0.3794359 RMSE and a 0.2687336 MAE. In words, worse approximations of the target distribution of the LWF are actually best at replicating the proxy for the SP500. A possible interpretation goes, once again, as that reported in the main text.}\
In the case of the EuroStoxx50 index, while the Bootstrap filter slightly increases both RMSE and MAE when N is raised to 50000, the APF shows mild improvements in its MAE performance, while the RMSE criterion points in the opposite direction. The criteria disagree also in the case of the LW filter, as the MAE mildly prefers the performance of N=10000, while the RMSE detects minor improvements for N=50000.
